Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-10-14 Manuscript/00-main.Rmd
Revise Manuscript

tar_load(WLM_train_dat)
# tar_load(tbl_one)
tar_load(ECB_bamlasso_cv)
tar_load(ECB_SBGAM_cv)
tar_load(WLM_bamlasso_cv)
tar_load(WLM_SBGAM_cv)
tar_load(tbl_real_time)
```

@@

```{r, results = "asis"}
tab_gaus$mgcv[4] <- "-"
tab_gaus$mgcv[5] <- "-"
tab_gaus %>% 
  mutate(p = as.integer(p)) %>% 
  rename(
    "P" = p,
    "BHAM-IWLS" = bglm_spline_de,
    "BHAM-CD" = blasso_spline,
    "COSSO" = cosso,
    "Adaptive COSSO" = acosso,
    "SB-GAM" = SB_GAM
  ) %>% 
  xtable(
    align = "cccccccc",
    caption = "The average and standard deviation of the out-of-sample $R^2$ measure for Gaussian outcomes over 50 iterations",
    label = "tab:gaus") %>% 
  print(comment = FALSE, include.rownames = FALSE)
  # flextable(
  #   # col_keys = c("P", "BHAM-IWLS", "BHAM-CD", "COSSO", "Adaptive COSSO", "mgcv", "SB-GAM")
  # ) %>% 
  # flextable::autofit() %>% 
  # flextable::set_caption(caption = "The average and standard deviation of the out-of-sample $R^2$ measure for Gaussian outcomes over 50 iterations.\\label{tab:gaus}")
```

@@

```{r, results = "asis"}
tab_binom$mgcv[4] <- "-"
tab_binom$mgcv[5] <- "-"
tab_binom %>% 
  mutate(p = as.integer(p)) %>% 
  rename(
    "P" = p,
    "BHAM-IWLS" = bglm_spline_de,
    "BHAM-CD" = blasso_spline,
    "COSSO" = cosso,
    "Adaptive COSSO" = acosso,
    "SB-GAM" = SB_GAM
  ) %>%  
  xtable(caption = "The average and standard deviation of the out-of-sample area under the curve measure for binomial outcomes over 50 iterations.", label = "tab:bin_auc") %>% 
  print(comment = FALSE, include.rownames = FALSE)
```

@@

```{r results = "asis"}
rbind(
  'BHAM-SSL' = ECB_bamlasso_cv %>% slice_max(auc) %>% select(-s0),
  'SB-GAM' = ECB_SBGAM_cv %>%  slice_max(auc) %>% select(-lambda)
) %>% 
  rownames_to_column("Methods") %>%
  select(-mae) %>% 
  rename(Deviance = deviance,
         AUC = auc,
         Brier = mse,
         Misclass = misclassification) %>% 
  xtable(
    caption = "Prediction performance of optimal BHAM and SB-GAM models for Emory Cardiovascular Biobank by 10-fold cross-validation, including deviance, area under the curve (AUC), Brier score, and misclassification error (Misclass) where class labels are defined using threshold = 0.5.",
    label = "tab:ECB_res"
  ) %>% 
  print(comment = FALSE, include.rownames = FALSE#, caption.placement = "top"
        )
```

@@

```{r results = "asis"}
rbind(
  'BHAM-SSL' = WLM_bamlasso_cv %>% slice_min(mse) %>% select(-s0),
  'SB-GAM' = WLM_SBGAM_cv %>%  slice_min(mse) %>% select(-lambda)
) %>% 
  rownames_to_column("Methods") %>%
  rename(
    Deviance = deviance,
    MSE = mse,
    MAE = mae,
    `$R^2$` = R2
  ) %>% 
    xtable(
    caption = "Prediction performance of of optimal BHAM and SB-GAM models for Weight Loss Maintenance Cohort by 10-fold cross-validation, including deviance, $R^2$,  mean squared error (MSE), and mean absolute error (MAE).",
    label = "tab:WLM_res"
  ) %>%
  print(comment = FALSE, include.rownames = FALSE, sanitize.text.function=function(x){x}#, caption.placement = "top"
        )
  # flextable() %>% 
  # flextable::autofit() %>% 
  # flextable::set_caption(caption = "Measures of optimal BHAM-SSL and SB-GAM models for Weight Loss Maintenance Cohort by 10-fold cross-validation\\label{tab:WLM_res}")
```

>>>>

tar_load(WLM_train_dat)
# tar_load(tbl_one)
tar_load(ECB_bamlasso_cv)
tar_load(ECB_SBGAM_cv)
tar_load(WLM_bamlasso_cv)
tar_load(WLM_SBGAM_cv)
tar_load(tbl_real_time)
tar_load(tab_time)
```

@@

```{r, results = "asis"}
tab_gaus$mgcv[4] <- "-"
tab_gaus$mgcv[5] <- "-"
tab_gaus %>% 
  mutate(p = as.integer(p)) %>% 
  rename(
    "P" = p,
    "BHAM-IWLS" = bglm_spline_de,
    "BHAM-CD" = blasso_spline,
    "COSSO" = cosso,
    "Adaptive COSSO" = acosso,
    "SB-GAM" = SB_GAM
  ) %>% 
  xtable(
    align = "cccccccc",
    caption = "The average and standard deviation of the out-of-sample $R^2$ measure for Gaussian outcomes over 50 iterations. The models of comparison include the proposed Bayesian hierarchical additive model (BHAM) fitted with Iterative Weighted Least Square (BHAM-IWLS) and Coordinate Descent (BHAM-CD) algorithms, component selection and smoothing operator (COSSO), adaptive COSSO, mgcv and sparse Bayesian generalized additive model (SB-GAM). mgcv doesn't provide estimation whe number of parameters exceeds sample size i.e. p = 100, 200.",
    label = "tab:gaus") %>% 
  print(comment = FALSE, include.rownames = FALSE)
  # flextable(
  #   # col_keys = c("P", "BHAM-IWLS", "BHAM-CD", "COSSO", "Adaptive COSSO", "mgcv", "SB-GAM")
  # ) %>% 
  # flextable::autofit() %>% 
  # flextable::set_caption(caption = "The average and standard deviation of the out-of-sample $R^2$ measures for Gaussian outcomes over 50 iterations. The models of comparison include the proposed Bayesian hierarchical additive model (BHAM) fitted with Iterative Weighted Least Square (BHAM-IWLS) and Coordinate Descent (BHAM-CD) algorithms, component selection and smoothing operator (COSSO), adaptive COSSO, mgcv and sparse Bayesian generalized additive model (SB-GAM). \\label{tab:gaus}")
```

@@

```{r, results = "asis"}
tab_binom$mgcv[4] <- "-"
tab_binom$mgcv[5] <- "-"
tab_binom %>% 
  mutate(p = as.integer(p)) %>% 
  rename(
    "P" = p,
    "BHAM-IWLS" = bglm_spline_de,
    "BHAM-CD" = blasso_spline,
    "COSSO" = cosso,
    "Adaptive COSSO" = acosso,
    "SB-GAM" = SB_GAM
  ) %>%  
  xtable(caption = "The average and standard deviation of the out-of-sample area under the curve measures for binomial outcomes over 50 iterations. The models of comparison include the proposed Bayesian hierarchical additive model (BHAM) fitted with Iterative Weighted Least Square (BHAM-IWLS) and Coordinate Descent (BHAM-CD) algorithms, component selection and smoothing operator (COSSO), adaptive COSSO, mgcv and sparse Bayesian generalized additive model (SB-GAM). mgcv doesn't provide estimation whe number of parameters exceeds sample size i.e. p = 100, 200.", label = "tab:bin_auc") %>% 
  print(comment = FALSE, include.rownames = FALSE)
```

\newpage
```{r}
tab_time %>% 
  select(dist, p, `BHAM-IWLS`,`BHAM-CD`, cosso, acosso, mgcv, `SB-GAM`) %>% 
  rename(Distribution = dist,
         P = p,
         COSSO = cosso,
         `Adaptive COSSO` = acosso) %>%

  flextable::flextable() %>% 
    merge_v(j = 1) %>% 
  flextable::set_caption(caption = "The average and standard deviation of computation time in seconds, including cross-validation and final model fitting, over 50 iterations. The models of comparison include the proposed Bayesian hierarchical additive model (BHAM) fitted with Iterative Weighted Least Square (BHAM-IWLS) and Coordinate Descent (BHAM-CD) algorithms, component selection and smoothing operator (COSSO), adaptive COSSO, mgcv and sparse Bayesian generalized additive model (SB-GAM). mgcv doesn't provide estimation whe number of parameters exceeds sample size i.e. p = 100, 200.\\label{tab:time_sim}")
```

@@

```{r results = "asis"}
rbind(
  'BHAM-CD' = ECB_bamlasso_cv %>% slice_max(auc) %>% select(-s0),
  'SB-GAM' = ECB_SBGAM_cv %>%  slice_max(auc) %>% select(-lambda)
) %>% 
  rownames_to_column("Methods") %>%
  select(-mae) %>% 
  rename(Deviance = deviance,
         AUC = auc,
         Brier = mse,
         Misclass = misclassification) %>% 
  xtable(
    caption = "Prediction performance of BHAM fitted with Cooridnate Descent algorithm (BHAM-CD) and SB-GAM models for Emory Cardiovascular Biobank by 10-fold cross-validation, including deviance, area under the curve (AUC), Brier score, and misclassification error (Misclass) where class labels are defined using threshold = 0.5.",
    label = "tab:ECB_res"
  ) %>% 
  print(comment = FALSE, include.rownames = FALSE#, caption.placement = "top"
        )
```

@@

```{r results = "asis"}
rbind(
  'BHAM-CD' = WLM_bamlasso_cv %>% slice_min(mse) %>% select(-s0),
  'SB-GAM' = WLM_SBGAM_cv %>%  slice_min(mse) %>% select(-lambda)
) %>% 
  rownames_to_column("Methods") %>%
  rename(
    Deviance = deviance,
    MSE = mse,
    MAE = mae,
    `$R^2$` = R2
  ) %>% 
    xtable(
    caption = "Prediction performance of of BHAM fitted with Cooridnate Descent algorithm (BHAM-CD)  and SB-GAM models for Weight Loss Maintenance Cohort by 10-fold cross-validation, including deviance, $R^2$,  mean squared error (MSE), and mean absolute error (MAE).",
    label = "tab:WLM_res"
  ) %>%
  print(comment = FALSE, include.rownames = FALSE, sanitize.text.function=function(x){x}#, caption.placement = "top"
        )
  # flextable() %>% 
  # flextable::autofit() %>% 
  # flextable::set_caption(caption = "Measures of optimal BHAM-SSL and SB-GAM models for Weight Loss Maintenance Cohort by 10-fold cross-validation\\label{tab:WLM_res}")
```

### 2021-10-14 Manuscript/02-Method.Rmd
Revise Manuscript

Following the GLM notation introduced in Section \ref{sec:intro}, we have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\bs x) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \bs \beta_j^T \bs X_j),
\end{equation}
and the data distribution is
\begin{equation}
f(\bs Y = \bs y| \bs \beta, \phi) = \prod\limits^n_{i=1}f( Y = y_i|\bs \beta, \phi),\nonumber
\end{equation}
with smoothing functions $B_j(x_j)$ of the variable $x_j, j = 1, \dots, p.$ The basis function matrix, i.e. the design matrix derived from the smoothing function $B_j(x_j)$, is denoted $\bs X_j$ for the variable $x_j$. The dimension of the design matrix depends on the choice of the smoothing function. For example, a cubic spline function with $k$ knots produces a $k+3$ elements column vector. Its corresponding smoothing penalty matrix is denoted as $\bs S_j$. $\bs \beta_j$ denotes the basis function coefficients for the $j$th variable such that $B_j(x_j) = \bs \beta_j^T \bs X_j$. With slight abuse of notation, we denote vectors and matrices in bold fonts $\bs \beta, \bs X$ with conformable dimensions, where scalar and random variables are denoted in with unbold fonts $\beta, X$. The matrix transposing operation is denoted with a superscript $^T$. To note, the proposed model can include parametric forms of variables in the model, treating the parametric function as a special case of the smoothing function, e.g. $B_j(x_j) = x_j$ with the smoothing penalty matrix defined as $\bs S_j = \begin{bmatrix}0\end{bmatrix}$.

@@

The family of spike-and-slab (SS) priors regression models dominates Bayesian high-dimensional analysis for its utility in outcome prediction and variable selection. We defer to Bai et al. [@Bai2021Review] for an in-depth introduction to spike-and-slab priors. To summarize, spike-and-slab priors are a family of mixture distributions that comprises a skinny spike density $f_{\text{spike}}(\cdot)$ for weak signals and a flat slab density $f_{\text{slab}}(\cdot)$ for strong signals, mathematically
$$
 \beta|\gamma \sim (1-\gamma)f_{\text{spike}}(\beta) + \gamma f_{\text{slab}}(\beta).
$$
The most distinct feature of SS priors is that it is conditioned on a latent binary variable $\gamma \in \{0,1\}$ that indicates whether the variable $x$ is included in the model. There are various spike-and-slab priors depending on the choice for the spike density $f_{\text{spike}}(\cdot)$ and the slab density $f_{\text{slab}}(\cdot)$, see George and McCulloch [@George1993; @George1997]; Chipman [@Chipman1996] for grouped variables; Brown et al. [@Brown1998] for multivariate outcomes; Ishwaran and Rao [@Ishwaran2005]; Clyde and George [@Clyde2004] and reference therein.

@@

The major criticism of early spike-and-slab models is being computationally prohibitive. [@Bai2021Review] Since then, many studies focus on alleviating the computational burden that sampling algorithms bear, which include EMVS [@Rockova2014a] and spike-and-slab Lasso [@Rockova2018b; @Rockova2018]. Particularly, the development of spike-and-slab Lasso  substantially improves the scalability of SS models, setting up the theoretical foundation for generalized models in -omics data analysis [@Tang2017a; @Tang2017; @Tang2018; @Tang2019]. The spike-and-slab Lasso (SSL) prior is composed of two double exponential distributions with mean 0 and different dispersion parameters, $0 < s_0 < s_1$, mathematically,
\begin{equation} 
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber
\end{equation}
Given that both double exponential distributions have a mean of 0 and the latent indicator $\gamma$ can only take the value of 0 or 1, the mixture double exponential distribution can be formulated as one single double exponential density,
\begin{equation} \label{eq:ssl}
\beta | \gamma \sim DE(0, S), 0 < s_0 < s_1,
\end{equation}
with the scale parameter $S = (1-\gamma)s_0 + \gamma s_1$. The SSL also mitigates the problem of EMVS where the weak signals are not shrink to zero, and hence is preferred in high-dimensional data analysis. We notice that Bai [@Bai2021] is the first to apply spike-and-slab lasso prior in the GAM framework, where the densities of the spike and slab components take the group lasso density [@Xu2015]. Bai [@Bai2021] takes an "all-in-all-out" strategy for function selection.

@@

In this section, we extend the two EM-based algorithms, EM-CD and EM-IWLS algorithms, to fit BHAMs. To note, the two proposed algorithms provides different utilities. The EM-CD algorithm is specifically for fitting BHAM-SSL with an expedited performance, recommending to use in high and ultra-high dimensional setting. A specific concern of EM-CD algorithm is that it provides no information for inference. In contrast, the EM-IWLS can extract the variance-covariance matrix and be used to construct Wald type inference. Moreover, the EM-IWLS is a more general model fitting algorithm that can be used for fitting not only SSL and continuous SS priors but also Student's t-priors and double exponential priors. To note, SB-GAM[@Bai2020; @Bai2021] also used an EM-CD algorithm. The main difference between proposed EM-CD algorithm and that in SB-GAM is that SB-GAM uses a blocked CD algorithm for their group prior, while the proposed prior is pairwise independent requiring no special treatment in the CD algorithm.

@@

For BHAMs, we define the parameters of interest as $\Theta = \{\bs \beta, \bs \theta, \phi\}$ and consider the latent binary indicators $\bs \gamma$ as nuisance parameters of the model, in other words the "missing" data. Our objective is to find the parameters $\Theta$ that maximize the posterior density function, or equivalently, the logarithm of the density function, 
$$
\begin{aligned}
& \text{argmax}_{\Theta}
\log f(\Theta, \bs \gamma| \textbf{y}, \textbf{X}) \\
&= \log f(\textbf{y}|\bs \beta, \phi) + \sum\limits_{j=1}^p\left[\log f(\beta_j|\gamma_j)+\sum\limits_{k=1}^{K_j} \log f(\beta^{\tp}_{jk}|\gamma^{\tp}_{j})\right]\\
& +\sum\limits_{j=1}^{p} \left[ (\gamma_j+\gamma_{j}^{\tp})\log \theta_j + (2-\gamma_j-\gamma_{j}^{\tp}) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log f(\theta_j),
\end{aligned}
$$  
where $f(\textbf{y}|\bs \beta, \phi)$ is the data distribution and $f(\theta)$ is the Beta(1,1) density.

@@

With $\bs z\sim \text{MVN}(\bs X \bs \beta, \phi \bs \Sigma)$ and $\bs \beta \sim \text{MVN}(0, \phi \bs \Sigma_{\tau^2})$, we can augment the two multivariate normal distributions and update the estimates for $\bs \beta$ and $\phi$ via least squares in each iteration of the EM algorithm. We create the augmented response, augmented data, and augmented variance-covariance matrix following \begin{align*}
& \bs z_* = \begin{bmatrix} \bs z\\ \bs 0\end{bmatrix} &&
  \bs X_* = \begin{bmatrix} \bs X \\ \bs I \end{bmatrix} &&
  \bs \Sigma_* = \begin{bmatrix} \bs \Sigma & \bs 0  \\ \bs 0 & \bs \Sigma_{\tau^2}/\phi \end{bmatrix}, &
\end{align*}
 such that 
$$
\bs z_* \sim \text{MVN}(\bs X_* \bs \beta , \phi \Sigma_*).
$$
Using the least squares estimators to update $\bs\beta$ and $\phi$, we have 
\begin{align*}
& \bs \beta^{(t)} = (\bs X_*^T \bs \Sigma^{-1} \bs X_*)^{-1}\bs X_*^T \bs \Sigma^{-1} \bs z_* && \phi^{(t)} = \frac{1}{n}(\bs z_*-X_*\bs \beta^{(t)})^T\bs \Sigma^{-1}(\bs z_*-X_*\bs \beta^{(t)}).&
\end{align*}
To note, the variance-covariance matrixof the coefficient estimates variance-covariance matrix can be derived in the EM-IWLS algorithm and in turn can be used for statistical inferences,
$$
  \text{Var}(\bs\beta^{(t)}) = (\bs X_*^T\bs \Sigma^{-1} \bs X_*)^{-1}\phi^{(t)}.
$$
The rest of the algorithm remains the same as described in Section \ref{sec:EMCD}.

>>>>

Following the GLM notation introduced in Section \ref{sec:intro}, we have a generalized additive model with link function $g(\cdot)$ and linear predictor
\begin{equation}\label{eq:gam}
\eta = \beta_0 + \sum\limits^p_{j=1}B_j(x_j) = \beta_0 + \sum\limits^p_{j=1} \bs \beta_j^T \bs X_j,
\end{equation}
with smoothing functions $B_j(x_j)$ of the variable $x_j, j = 1, \dots, p.$ The outcome $Y$ follows a exponential family distribution with density function $f(y)$, mean $\mu = g^{-1}(\eta)$ and dispersion parameter $\phi$, and the data distribution is
\begin{equation}
f(\bs Y = \bs y| \bs \beta, \phi) = \prod\limits^n_{i=1}f( Y = y_i|\bs \beta, \phi),\nonumber
\end{equation}
 The basis function matrix, i.e. the design matrix derived from the smoothing function $B_j(x_j)$, is denoted $\bs X_j$ for the variable $x_j$. The dimension of the design matrix depends on the choice of the smoothing function. For example, a cubic spline function with $k$ knots produces a $k+3$ elements column vector. Its corresponding smoothing penalty matrix is denoted as $\bs S_j$. $\bs \beta_j$ denotes the basis function coefficients for the $j$th variable such that $B_j(x_j) = \bs \beta_j^T \bs X_j$. With slight abuse of notation, we denote vectors and matrices in bold fonts $\bs \beta, \bs X$ with conformable dimensions, where scalar and random variables are denoted in with unbold fonts $\beta, X$. The matrix transposing operation is denoted with a superscript $^T$. To note, the proposed model can include parametric forms of variables in the model, treating the parametric function as a special case of the smoothing function, e.g. $B_j(x_j) = x_j$ with the smoothing penalty matrix defined as $\bs S_j = \begin{bmatrix}0\end{bmatrix}$.

@@

The family of spike-and-slab (SS) priors regression models is one of most commonly used models in high-dimensional data analysis for its utility in outcome prediction and variable selection. We defer to Bai et al. [@Bai2021Review] for an in-depth introduction to spike-and-slab priors. To summarize, spike-and-slab priors are a family of mixture distributions that comprises a skinny spike density $f_{\text{spike}}(\cdot)$ for weak signals and a flat slab density $f_{\text{slab}}(\cdot)$ for strong signals, mathematically
$$
 \beta|\gamma \sim (1-\gamma)f_{\text{spike}}(\beta) + \gamma f_{\text{slab}}(\beta).
$$
The most distinct feature of SS priors is that it is conditioned on a latent binary variable $\gamma \in \{0,1\}$ that indicates whether the variable $x$ is included in the model. There are various spike-and-slab priors depending on the choice for the spike density $f_{\text{spike}}(\cdot)$ and the slab density $f_{\text{slab}}(\cdot)$, see George and McCulloch [@George1993; @George1997]; Chipman [@Chipman1996] for grouped variables; Brown et al. [@Brown1998] for multivariate outcomes; Ishwaran and Rao [@Ishwaran2005]; Clyde and George [@Clyde2004] and reference therein. Two of most popular spike and slab priors are spike-and-slab normal prior [@George1993] and spike-and-slab double exponential prior[@Rockova2018].

@@

The major criticism of early spike-and-slab models is being computationally prohibitive. [@Bai2021Review] Since then, many studies focus on alleviating the computational burden that sampling algorithms bear, which include EMVS based on spike-and-slab normal prior[@Rockova2014a] and spike-and-slab Lasso [@Rockova2018b; @Rockova2018]. Particularly, the development of spike-and-slab Lasso  substantially improves the scalability of SS models, setting up the theoretical foundation for generalized models in -omics data analysis [@Tang2017a; @Tang2017; @Tang2018; @Tang2019]. The spike-and-slab Lasso (SSL) prior is composed of two double exponential distributions with mean 0 and different dispersion parameters, $0 < s_0 < s_1$, mathematically,
\begin{equation} 
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber
\end{equation}
Given that both double exponential distributions have a mean of 0 and the latent indicator $\gamma$ can only take the value of 0 or 1, the mixture double exponential distribution can be formulated as one single double exponential density,
\begin{equation} \label{eq:ssl}
\beta | \gamma \sim DE(0, S), 0 < s_0 < s_1,
\end{equation}
with the scale parameter $S = (1-\gamma)s_0 + \gamma s_1$. The SSL also mitigates the problem of EMVS where the weak signals are not shrink to zero, and hence is preferred in high-dimensional data analysis. We notice that Bai [@Bai2021] is the first to apply spike-and-slab lasso prior in the GAM framework, where the densities of the spike and slab components take the group lasso density [@Xu2015]. Bai [@Bai2021] takes an "all-in-all-out" strategy for function selection.

@@

In this section, we extend the two EM-based algorithms, EM-CD and EM-IWLS algorithms, to fit BHAMs. To note, the two proposed algorithms provides different utilities. The EM-CD algorithm is specifically for fitting BHAM-SSL with an expedited performance, recommending to use in high and ultra-high dimensional setting. A specific concern of EM-CD algorithm is that it provides no information for inference. In contrast, the EM-IWLS can estimate the variance-covariance matrix of the coefficients. Moreover, the EM-IWLS is a more general model fitting algorithm that can be used for fitting not only SSL and continuous SS priors but also Student's t-priors and double exponential priors. To note, SB-GAM[@Bai2020; @Bai2021] also used an EM-CD algorithm. The main difference between proposed EM-CD algorithm and that in SB-GAM is that SB-GAM uses a blocked CD algorithm for their group prior, while the proposed prior is pairwise independent requiring no special treatment in the CD algorithm.

@@

For BHAMs, we define the parameters of interest as $\Theta = \{\bs \beta, \bs \theta, \phi\}$ and consider the latent binary indicators $\bs \gamma$ as nuisance parameters of the model, in other words the "missing" data. Our objective is to find the parameters $\Theta$ that maximize the posterior density function, or equivalently, the logarithm of the density function, 
$$
\begin{aligned}
& \text{argmax}_{\Theta}
\log f(\Theta, \bs \gamma| \textbf{y}, \textbf{X}) \\
&= \log f(\textbf{y}|\bs \beta, \phi) + \sum\limits_{j=1}^p\left[\log f(\beta_j|\gamma_j)+\sum\limits_{k=1}^{K_j} \log f(\beta^{\tp}_{jk}|\gamma^{\tp}_{j})\right]\\
& +\sum\limits_{j=1}^{p} \left[ (\gamma_j+\gamma_{j}^{\tp})\log \theta_j + (2-\gamma_j-\gamma_{j}^{\tp}) \log (1-\theta_j)\right] +  \sum\limits_{j=1}^{p}\log f(\theta_j),
\end{aligned}
$$  
where $f(\textbf{y}|\bs \beta, \phi)$ is the data distribution and $f(\theta)$ is the Beta(1,1) density. We choose non-informative prior for the intercept $\beta_0$ and the dispersion parameter $\phi$; for example, $f(\beta_0|\tau_0^2)=N(0,\tau_0^2)$ with $\tau^2_0$ set to a large value and $f(\log \phi) \propto 1$.

@@

With $\bs z\sim \text{MVN}(\bs X \bs \beta, \phi \bs \Sigma)$ and $\bs \beta \sim \text{MVN}(0, \phi \bs \Sigma_{\tau^2})$, we can augment the two multivariate normal distributions and update the estimates for $\bs \beta$ and $\phi$ via least squares in each iteration of the EM algorithm. We create the augmented response, augmented data, and augmented variance-covariance matrix following \begin{align*}
& \bs z_* = \begin{bmatrix} \bs z\\ \bs 0\end{bmatrix} &&
  \bs X_* = \begin{bmatrix} \bs X \\ \bs I \end{bmatrix} &&
  \bs \Sigma_* = \begin{bmatrix} \bs \Sigma & \bs 0  \\ \bs 0 & \bs \Sigma_{\tau^2}/\phi \end{bmatrix}, &
\end{align*}
 such that 
$$
\bs z_* \sim \text{MVN}(\bs X_* \bs \beta , \phi \Sigma_*).
$$
Using the least squares estimators to update $\bs\beta$ and $\phi$, we have 
\begin{align*}
& \bs \beta^{(t)} = (\bs X_*^T \bs \Sigma^{-1} \bs X_*)^{-1}\bs X_*^T \bs \Sigma^{-1} \bs z_* && \phi^{(t)} = \frac{1}{n}(\bs z_*-X_*\bs \beta^{(t)})^T\bs \Sigma^{-1}(\bs z_*-X_*\bs \beta^{(t)}).&
\end{align*}
To note, the variance-covariance matrixof the coefficient estimates variance-covariance matrix can be derived in the EM-IWLS algorithm and in turn can be used for statistical inferences,
$$
  \text{Var}(\bs\beta^{(t)}) = (\bs X_*^T\bs \Sigma^{-1} \bs X_*)^{-1}\phi^{(t)}.
$$

Totally, the proposed EM-iterative weighted least square algorithm is summarized as follows:

    E-step: calculate $E(\gamma_{j})$, $E(\gamma^\tp_{j})$ and $E(\tau^{-2}_{j})$, $E({\tau^\tp}^{-2}_{jk})$ with estimates of $\Theta^{(t-1)}$ from the previous iteration

    a) Based on the current value of $\beta$, calculate the pseudo-data $z_i^{(t)}$ and the pseudo-weights $w_i^{(t)}$
    b) Update $\bs \beta^{(t)}$ by runing the augmented weighted least squared
    c) If $\phi$ is present, update $\phi$

Similar to EM-CD, we assess convergence by the criterion, $|d^{(t)}-d^{(t-1)}|/(0.1+|d^{(t)}|)<\epsilon$, where $\epsilon$ is a small value (say $10^{-5}$).

### 2021-10-14 Manuscript/04-Real_Data.Rmd
Revise Manuscript

## Emory Cardiovascular Biobank
We use the proposed models BHAM-SSL to analyze a metabolic dataset from a recently published research [@Mehta2020] studying plasma metabolomic profile on the three-year all-cause mortality among patients undergoing cardiac catheterization. The dataset is publicly available via \textit{Dryad} [@Mehta2020_data]. It contains in total of 776 subjects from two cohorts. As there is a large number of non-overlapping features among the two cohorts, we use the cohort with larger sample size (N=454). There were initially 6796 features in the dataset, which is too large to be practically meaningful to analyze. Hence, we performed a univariate screening procedure on the features, via GAM implemented in \texttt{mgcv}, and chose the the top 200 features with smallest p-values. We used 5-knot spline additive models for binary outcome using two different models, the proposed BHAM-SSL and the SB-GAM. 10-Fold cross-validation(CV) are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples were used for prediction performance evaluation, where deviance, AUC, Brier score defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and misclassication error defined as $\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)$ were calculated and tabulated in Table \ref{tab:ECB_res}.
\begin{center}
Table \ref{tab:ECB_res} here
\end{center}

## Weight Loss Maintenance Cohort 
We use the proposed models BHAM-SSL to analyze metabolomics data from a recently published study [@Bihlmeyer2021] on the association between metabolic biomarkers and weight loss, where the dataset is publicly available [@Bihlmeyer2021_data]. In this analysis, we primarily focused on the analysis of one of the three studies included, weight loss maintenance (WLM) cohort [@Svetkey2008], due to the drastically different intervention effects. In the dataset, 765 metaboliltes in baseline plasma collected were profiled using liquid chromatography mass spectrometry. Quality control and natural log transformation were performed during metabolites data preparation. The outcome of interest was standardized percent change in insulin resistance, and hence modeled using a Gaussian model. After removing missing datapoints and addressing outliers in the data, there were $p$=`r nrow(WLM_train_dat)` features remaining in the analysis. 5-Knot spline additive models for the gaussian outcome are constructed using two different models, the proposed BHAM-SSL and the SB-GAM. 10-Fold cross-validation(CV) are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples were used for prediction performance evaluation, where deviance, mean squared error (MSE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and mean absolute error (MAE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}|y_i - \hat y_i|$ were calculated and tabulated in Table \ref{tab:WLM_res}.
\begin{center}
Table \ref{tab:WLM_res} here
\end{center}

>>>>

## Emory Cardiovascular Biobank
We use the proposed models BHAM-SSL to analyze a metabolic dataset from a recently published research [@Mehta2020] studying plasma metabolomic profile on the three-year all-cause mortality among patients undergoing cardiac catheterization. The dataset is publicly available via \textit{Dryad} [@Mehta2020_data]. It contains in total of 776 subjects from two cohorts. As there is a large number of non-overlapping features among the two cohorts, we use the cohort with larger sample size (N=454). There were initially 6796 features in the dataset, which is too large to be practically meaningful to analyze. Hence, we performed a univariate screening procedure on the features, via GAM implemented in \texttt{mgcv}, and chose the the top 200 features with smallest p-values. We used 5-knot spline additive models for binary outcome using two different models, the proposed BHAM-SSL and the SB-GAM. 10-Fold cross-validation(CV) are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples were used for prediction performance evaluation, where deviance, AUC, Brier score defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and misclassication error defined as $\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)$ were calculated. BHAM-CD obtained superior AUC, Brier score, and misclassification error in the out-of-bag samples compared to SB-GAM (Table \ref{tab:ECB_res}).
\begin{center}
Table \ref{tab:ECB_res} here
\end{center}

## Weight Loss Maintenance Cohort 
We use the proposed models BHAM-SSL to analyze metabolomics data from a recently published study [@Bihlmeyer2021] on the association between metabolic biomarkers and weight loss, where the dataset is publicly available [@Bihlmeyer2021_data]. In this analysis, we primarily focused on the analysis of one of the three studies included, weight loss maintenance (WLM) cohort [@Svetkey2008], due to the drastically different intervention effects. In the dataset, 765 metaboliltes in baseline plasma collected were profiled using liquid chromatography mass spectrometry. Quality control and natural log transformation were performed during metabolites data preparation. The outcome of interest was standardized percent change in insulin resistance, and hence modeled using a Gaussian model. After removing missing datapoints and addressing outliers in the data, there were $p$=`r nrow(WLM_train_dat)` features remaining in the analysis. 5-Knot spline additive models for the gaussian outcome are constructed using two different models, the proposed BHAM-SSL and the SB-GAM. 10-Fold cross-validation(CV) are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples were used for prediction performance evaluation, where deviance, $R^2$, mean squared error (MSE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and mean absolute error (MAE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}|y_i - \hat y_i|$ were calculated. BHAM-CD obtained superior R2, MSE, and MAE in the out-of-bag samples compared to SB-GAM (Table \ref{tab:WLM_res}).
\begin{center}
Table \ref{tab:WLM_res} here
\end{center}
